Rhyme: A Query Language for Nested Data Structures

### 2024-12-14 paper.tex
section 4 intro finish

\section{Introduction}
\iffalse
Many languages exist for computing high dimensional nested data structures like tabular data, JSON objects and tensors. Many researchers use these languages to perform data retrieval and computation together in fields like scientific computation and machine learning. Sparse tensor algebra have also been widely used in both neural networks and scientific simulations to efficiently store and compute sparse data. However, relational algebra and tensor computation often requires different programming languages and they may store data in different formats. So the developers need to write different tasks in different languages and integrates them together through text/binary files. This is both time costly and make the project difficult to maintain. \par
\fi

@@

\item \textbf{Generators}: Generator types are inferred based on the type schema of the data sources they iterate over. As mentioned earlier, the generator's type is the intersection of the key types of its data sources. For instance, if \inline{input} is a 1-D dense tensor with \inline{i32} elements and the generator \inline{*} is applied, the type of \inline{*} is \inline{u32}, and the type of \inline{input.*} is \inline{i32}. Since \inline{*} iterates over a continuous range, it can be incremented by one for each iteration. However, this is not guaranteed for sparse tensors. \end{itemize}\par
\iffalse
\rhyme 's type system lays the foundations for specialized code generation for strongly typed languages like C/C++ as well as sparse tensor algebra. Dense tensor and sparse tensor has different characteristics and require different methods for both data loading as well as efficient co-iteration. We will cover how this type system facilitates efficient code generation for Sparse Tensor Algebra in the next section.
\fi
\rhyme's type system forms the foundation for specialized code generation for strongly typed languages like C/C++ as well as for sparse tensor algebra. Dense tensors and sparse tensors have distinct characteristics, requiring different strategies for both data loading and efficient co-iteration. In the next section, we will discuss how this type system supports the efficient code generation of sparse tensor algebra.
\section{Sparse Tensor Algebra}\label{sparse}
\section{Evaluation}\label{evaluation}
\section{Conclusion}\label{conclusion}
%\newpage
\bibliographystyle{acm}
\bibliography{paper}
\end{document}
\endinput

>>>>

\section{Introduction}\label{introduction}
\iffalse
Many languages exist for computing high dimensional nested data structures like tabular data, JSON objects and tensors. Many researchers use these languages to perform data retrieval and computation together in fields like scientific computation and machine learning. Sparse tensor algebra have also been widely used in both neural networks and scientific simulations to efficiently store and compute sparse data. However, relational algebra and tensor computation often requires different programming languages and they may store data in different formats. So the developers need to write different tasks in different languages and integrates them together through text/binary files. This is both time costly and make the project difficult to maintain. \par
\fi

@@

\item \textbf{Generators}: Generator types are inferred based on the type schema of the data sources they iterate over. As mentioned earlier, the generator's type is the intersection of the key types of its data sources. For instance, if \inline{input} is a 1-D dense tensor with \inline{i32} elements and the generator \inline{*} is applied, the type of \inline{*} is \inline{u32}, and the type of \inline{input.*} is \inline{i32}. Since \inline{*} iterates over a continuous range, it can be incremented by one for each iteration. However, this is not guaranteed for sparse tensors. \end{itemize}\par
\rhyme 's type system lays the foundations for specialized code generation for strongly typed languages like C/C++ as well as sparse tensor algebra. Dense tensor and sparse tensor has different characteristics and require different methods for both data loading as well as efficient co-iteration. We will cover how this type system facilitates efficient code generation for Sparse Tensor Algebra in the next section.
\rhyme's type system forms the foundation for specialized code generation for strongly typed languages like C/C++ as well as for sparse tensor algebra. Dense tensors and sparse tensors have distinct characteristics, requiring different strategies for both data loading and efficient co-iteration. In the next section, we will discuss how this type system supports the efficient code generation of sparse tensor algebra.
\section{Sparse Tensor Algebra}\label{sparse}
In this section, we will discuss how \rhyme supports specialized code generation for sparse tensor algebra and how it manage to efficiently co-iterate over sparse tensors inside a multi-paradigm framework.\par
As we mentioned in Section \ref{introduction}, sparse tensor is widely used in both scientific computation and machine learning applications. Existing approach either uses hand written kernels to execute sparse tensor algebra or develops a sparse tensor compiler to automatically generate code given the tensor computation notations. However, integrating efficient sparse tensor algebra with relational algebra into a declarative language framework still remains a challenging problem. \rhyme manages to support efficient code generation for sparse tensor algebra inside a unified multi-paradigm framework.\par
Sparse tensor code generation reuses the same IR and dependency analysis pass as other domains, as the data and control dependencies are the same as dense tensor computation. After the user provides type schema indicating that the input is a sparse tensor, we will perform type inference and checking to assign valid types to all sub-expressions and intermediate temporary variables. We will also generate specialized functions to read/load the sparse tensor into memory as the sparse tensors are different format than dense formats. We support compressed notations of sparse vectors, where the non-zero values and their corresponding indexes are stored. Due to the sparsity nature of the data, sparse data does not offer constant-time look-up as the indexes stored are not continuous, a binary search will need to be taken to check existence of an index. Although sparse tensors does not offer constant-time random-access, it reduces the memory storage by only storing the non-fill (non-zero) elements. This compressed format also offers the opportunity for a more efficient way of computing tensors as the computation on zero elements can be skipped and we only need to compute for non-zero elements. For example, if we compute a dot product of two sparse vectors, we only need to multiply the indexes that all non-zero in both vectors.\par
In this section, we discuss how \rhyme supports specialized code generation for sparse tensor algebra and how it efficiently co-iterates over sparse tensors within a multi-paradigm framework.\par
As mentioned in Section \ref{introduction}, sparse tensors are widely used in both scientific computation and machine learning applications. Existing approaches either rely on hand-written kernels for executing sparse tensor algebra or develop sparse tensor compilers to automatically generate code from tensor computation notations. However, integrating efficient sparse tensor algebra with relational algebra into a declarative language framework remains a significant challenge. \rhyme addresses this challenge by enabling efficient code generation for sparse tensor algebra within a unified multi-paradigm framework.\par
Sparse tensor code generation in \rhyme reuses the same IR and dependency analysis passes as other domains, as the data and control dependencies for sparse tensors are identical to those for dense tensor computations. When the user provides a type schema indicating that the input is a sparse tensor, \rhyme performs type inference and checking to assign valid types to all sub-expressions and intermediate temporary variables. Additionally, \rhyme generates specialized functions to read and load sparse tensors into memory, as their storage format differs from dense tensor formats.\par
\rhyme supports compressed representations of sparse vectors, where only the non-zero values and their corresponding indices are stored. Due to the sparse nature of the data, sparse tensors do not offer constant-time look-up because the stored indices are not continuous. A binary search is required to check the existence of an index. While sparse tensors lack constant-time random access, they significantly reduce memory usage by storing only the non-fill (non-zero) elements. This compressed format also provides an opportunity for more efficient computations, as operations on zero elements can be skipped, and computations are performed only on non-zero elements. For example, when computing the dot product of two sparse vectors, it is sufficient to multiply the elements corresponding to indices that are non-zero in both vectors.
\subsection*{Efficient co-iteration of sparse vectors}
\subsection*{Implementation}
Different sparsity patterns require different ways of iterating tensors.
\section{Evaluation}\label{evaluation}
\section{Conclusion}\label{conclusion}
%\newpage
\bibliographystyle{acm}
\bibliography{paper}
\end{document}
\endinput
